---
title: "Interaction and effect size"
subtitle: "size matters"
author: "work in progress / enrico toffalini @ psicostat"
date: "February 21^st^, 2024"
output: html_document
---
\raggedright
\break\break

```{r setup-knitr, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### Initial setup

```{r setup-general} 
rm(list=ls())
library(ggplot2)
library(effects)
library(effsize)
rnd = function(x,digits=2){ return(format(round(x,digits),nsmall=digits)) }
load("interaction_datasets.RData")
tsz=25
```

Fully open code at: https://github.com/psicostat/workshops/tree/main/Interaction%20introduction%202024

----------------------------------------------------------------------------------------

----------------------------------------------------------------------------------------

----------------------------------------------------------------------------------------


# WHAT'S IN A "Cohen's *f*"?


Cohen's *f* is often used for expressing effect size in interactions. General formula for Cohen's *f* is like this:

$$Cohen's f = \sqrt{\frac{R^2}{1 - R^2}}$$

However, for interactions we should consider what is *added* by the interaction alone (above and beyond the main effects):

$$Cohen's f = \sqrt{\frac{R^2_1 - R^2_0}{1 - R^2_1}}$$

Cohen's *f* = 0.25 (about Cohen's *f* ^2^ = 0.06) is often taken as "medium" effect size. In fact, Cohen's *f* = 0.40 (about Cohen's *f* ^2^ = 0.15) is also taken as "medium" sometimes. Note that Cohen's *f* = 0.25 requires R^2^ = 0.06, corresponding to about *r* = 0.24, while Cohen's *f* = 0.40 requires R^2^ = 0.15, corresponding to about *r* = 0.39. **So, it makes more sense to say that Cohen's *f* = 0.25 is "medium".**

Anyways... what does that mean in actual interactions? The problem is that there are infinite cases that lead to the same effect size (e.g., Cohen's *f*) in interactions. 

All examples below will present about Cohen's *f* ≈ 0.25 (Cohen's *f* ^2^ ≈ 0.06). 


----------------------------------------------------------------------------------------

### Example: Fully crossed interaction (2 x 2)

May happen in well-designed experimental studies, but otherwise infrequent in real life (?).
Here, each main effect is small or even null, and the interaction accounts for most or all explained variance.

```{r figure-fully-crossed-interaction-2x2, cache=T, echo=FALSE, out.width="75%", fig.asp=0.7}
fit0 = lm(y~x+w,data=d22)
fit1 = lm(y~x*w,data=d22)
eff = data.frame(allEffects(fit1)$"x:w")
eff$letter = toupper(letters[1:nrow(eff)])
pd = position_dodge(width=.2)
ggplot(eff,aes(x=x,y=fit,group=w,color=w,linetype=w,shape=w))+
  geom_point(size=5,position=pd)+
  geom_line(linewidth=1,position=pd)+
  geom_errorbar(aes(ymin=lower,ymax=upper),linetype=1,width=.1,linewidth=1,position=pd)+
  theme(text=element_text(size=tsz))+
  ylab("y (outcome, z-score)")+
  geom_text(aes(x=as.numeric(x)+.05,y=fit+.1,label=letter),color="#888888",size=4.5,position=pd)

cf = cohens_f(fit1,fit0)
R2_0 = summary(fit0)$r.squared
R2_1 = summary(fit1)$r.squared
```

In model without interaction, R^2^ =  **``r rnd(R2_0)``**. In model with interaction, R^2^ =  **``r rnd(R2_1)``**.\
Overall, Cohen's *f* = **``r rnd(cf)``** (Cohen's *f* ^2^ = **``r rnd(cf^2)``**).

But... a simpler way to understand the above effect may be via Standardized Mean Differences (SMDs, such as Cohen's *d*): 

- the effect size (SMD) for X when W = 0 is Cohen's *d* = **``r rnd(cohen.d(d22$y[d22$w==0]~d22$x[d22$w==0])$estimate*(-1))``** (cell B - A).

- the effect size (SMD) for X when W = 1 is Cohen's *d* = **``r rnd(cohen.d(d22$y[d22$w==1]~d22$x[d22$w==1])$estimate*(-1))``** (cell D - C).


----------------------------------------------------------------------------------------

### Example: Treatment-like effect (2 x 2)

Imagine a situation like a treated-vs-control-group experiment with pre-test and post-test. In one condition (w = 0) x has not effect over y, while in the other (w = 1) x has a clearly visible effect over y. The interaction coefficients tells you (as usual) the difference-between-differences across conditions. 

```{r figure-treatment-like-interaction-2x2, cache=T, echo=FALSE, out.width="75%", fig.asp=0.7}
fit0 = lm(y~x+w,data=dt22)
fit1 = lm(y~x*w,data=dt22)
eff = data.frame(allEffects(fit1)$"x:w")
eff$letter = toupper(letters[1:nrow(eff)])
pd = position_dodge(width=.2)
ggplot(eff,aes(x=x,y=fit,group=w,color=w,linetype=w,shape=w))+
  geom_point(size=5,position=pd)+
  geom_line(linewidth=1,position=pd)+
  geom_errorbar(aes(ymin=lower,ymax=upper),linetype=1,width=.1,linewidth=1,position=pd)+
  theme(text=element_text(size=tsz))+
  ylab("y (outcome, z-score)")+
  geom_text(aes(x=as.numeric(x)+.05,y=fit+.1,label=letter),color="#888888",size=4.5,position=pd)

cf = cohens_f(fit1,fit0)
R2_0 = summary(fit0)$r.squared
R2_1 = summary(fit1)$r.squared
```

In model without interaction, R^2^ =  **``r rnd(R2_0)``**. In model with interaction, R^2^ =  **``r rnd(R2_1)``**.\
Overall, Cohen's *f* = **``r rnd(cf)``** (Cohen's *f* ^2^ = **``r rnd(cf^2)``**).

But note how much the effects must vary across conditions to get there:

- the effect size (SMD) for X when W = 0 is Cohen's *d* = **``r rnd(cohen.d(dt22$y[dt22$w==0]~dt22$x[dt22$w==0])$estimate*(-1))``** (cell B - A).

- the effect size (SMD) for X when W = 1 is Cohen's *d* = **``r rnd(cohen.d(dt22$y[dt22$w==1]~dt22$x[dt22$w==1])$estimate*(-1))``** (cell D - C).

...is this really a ***medium*** effect?! 😱


----------------------------------------------------------------------------------------

### Example: Treatment-like effect (2 x 3)

Imagine a situation like a treated-vs-control-group experiment with pre-test, post-test, and follow-up. Not much different from before, and it's a "familiar" situation. However, there are 2 interaction coefficients here; using the default contrasts, they tell you how the difference between levels of w (e.g., groups) varies from x = "00" to x = "01" (1^st^ interaction coefficient), and from x = "00" to x = "10" (2^nd^ interaction coefficient). 

```{r figure-treatment-like-interaction-2x3, cache=T, echo=FALSE, out.width="75%", fig.asp=0.7}
fit0 = lm(y~x+w,data=dt23)
fit1 = lm(y~x*w,data=dt23)
eff = data.frame(allEffects(fit1)$"x:w")
eff$letter = toupper(letters[1:nrow(eff)])
pd = position_dodge(width=.2)
ggplot(eff,aes(x=x,y=fit,group=w,color=w,linetype=w,shape=w))+
  geom_point(size=5,position=pd)+
  geom_line(linewidth=1,position=pd)+
  geom_errorbar(aes(ymin=lower,ymax=upper),linetype=1,width=.1,linewidth=1,position=pd)+
  theme(text=element_text(size=tsz))+
  ylab("y (outcome, z-score)")+
  geom_text(aes(x=as.numeric(x)+.06,y=fit+.12,label=letter),color="#888888",size=4.5,position=pd)

cf = cohens_f(fit1,fit0)
R2_0 = summary(fit0)$r.squared
R2_1 = summary(fit1)$r.squared
```

In model without interaction, R^2^ =  **``r rnd(R2_0)``**. In model with interaction, R^2^ =  **``r rnd(R2_1)``**.\
Overall, Cohen's *f* = **``r rnd(cf)``** (Cohen's *f* ^2^ = **``r rnd(cf^2)``**).

But note how much the effects must vary across conditions to get there:

- the effect size (SMD) at time x = "00" (cell D - A) is Cohen's *d* = **``r rnd(cohen.d(dt23$y[dt23$x=="00"]~dt23$w[dt23$x=="00"])$estimate*(-1))``**.

- the effect size (SMD) at time x = "01" (cell E - B) is Cohen's *d* = **``r rnd(cohen.d(dt23$y[dt23$x=="01"]~dt23$w[dt23$x=="01"])$estimate*(-1))``**.

- the effect size (SMD) at time x = "10" (cell F - C) is Cohen's *d* = **``r rnd(cohen.d(dt23$y[dt23$x=="10"]~dt23$w[dt23$x=="10"])$estimate*(-1))``**.

...is this really a ***medium*** effect?! 😱


----------------------------------------------------------------------------------------

### Example: Group moderate continuous effect (2 x continuous)

Imagine studying how a relationship between x and y is different across two populations. 

```{r figure-treatment-like-interaction-2xcontinuous, cache=T, echo=FALSE, out.width="75%", fig.asp=0.7}
fit0 = lm(y~x+w,data=d2c)
fit1 = lm(y~x*w,data=d2c)
eff = data.frame(allEffects(fit1,xlevels=list(x=seq(-2,2,.1)))$"x:w")
ggplot(eff,aes(x=x,y=fit,group=w,color=w,linetype=w,fill=w))+
  geom_ribbon(aes(ymin=lower,ymax=upper),linetype=1,linewidth=1,alpha=.2,color=NA)+
  geom_line(linewidth=1)+
  theme(text=element_text(size=tsz))+
  ylab("y (outcome, z-score)")

cf = cohens_f(fit1,fit0)
R2_0 = summary(fit0)$r.squared
R2_1 = summary(fit1)$r.squared
```

In model without interaction, R^2^ =  **``r rnd(R2_0)``**. In model with interaction, R^2^ =  **``r rnd(R2_1)``**.\
Overall, Cohen's *f* = **``r rnd(cf)``** (Cohen's *f* ^2^ = **``r rnd(cf^2)``**).

But note how much the effects must vary across conditions to get there:

- the x-y correlation in sample w = 0 is Pearson's *r* = **``r rnd(cor(d2c$x[d2c$w==0],d2c$y[d2c$w==0]))``**

- the x-y correlation in sample w = 0 is Pearson's *r* = **``r rnd(cor(d2c$x[d2c$w==1],d2c$y[d2c$w==1]))``**

...is this really a ***medium*** effect?! 😱


----------------------------------------------------------------------------------------

### Example: Continuous relationship moderated by another continuous variable (continuous x continuous)

Imagine studying how a relationship between x and y is different across different levels of a continuous moderator z. 

```{r figure-treatment-like-interaction-continuousxcontinuous, cache=T, echo=FALSE, out.width="75%", fig.asp=0.7}
fit0 = lm(y~x+w,data=dcc)
fit1 = lm(y~x*w,data=dcc)
eff = data.frame(allEffects(fit1,xlevels=list(x=seq(-2,2,.1),w=c(-1.5,-1,-0.5,0,0.5,1,1.5)))$"x:w")
eff$w = as.factor(eff$w)
ggplot(eff,aes(x=x,y=fit,group=w,color=w,fill=w))+
  geom_ribbon(aes(ymin=lower,ymax=upper),linetype=1,linewidth=1,alpha=.2,color=NA)+
  geom_line(linewidth=1)+
  theme(text=element_text(size=tsz))+
  ylab("y (outcome, z-score)")

cf = cohens_f(fit1,fit0)
R2_0 = summary(fit0)$r.squared
R2_1 = summary(fit1)$r.squared
```

In model without interaction, R^2^ =  **``r rnd(R2_0)``**. In model with interaction, R^2^ =  **``r rnd(R2_1)``**.\
Overall, Cohen's *f* = **``r rnd(cf)``** (Cohen's *f* ^2^ = **``r rnd(cf^2)``**).

The graphical representation above, however, is a little bit messy. We could simplify that by keeping just 3 lines at interpretable points (e.g., z-score of the moderator W equal -1, 0, +1):

```{r figure-treatment-like-interaction-continuousxcontinuous-simplified, cache=T, echo=FALSE, out.width="75%", fig.asp=0.7}
eff = data.frame(allEffects(fit1,xlevels=list(x=seq(-2,2,.1),w=c(-1,0,1)))$"x:w")
eff$w = as.factor(eff$w)
ggplot(eff,aes(x=x,y=fit,group=w,color=w,fill=w))+
  geom_ribbon(aes(ymin=lower,ymax=upper),linetype=1,linewidth=1,alpha=.2,color=NA)+
  geom_line(linewidth=1)+
  scale_color_manual(values=c("#C49A00","#00C094","#A58AFF"))+
  scale_fill_manual(values=c("#C49A00","#00C094","#A58AFF"))+
  theme(text=element_text(size=tsz))+
  ylab("y (outcome, z-score)")
#"#F8766D" "#C49A00" "#53B400" "#00C094" "#00B6EB" "#A58AFF" "#FB61D7"
```

Getting the estimates of the x-y relationship at given points of w, like in the previous case, can be done by looking at the coefficient "x" in the model summary, knowing that the "x:w" coefficient means how the effect of x on y varies for every unitary +1 increase of the moderator w. 

Let's have a look at the summary:

```{r}
summary(fit1)
```

In fact, y is not really on a z-score metrics here. Its residual SD that is about 1.0. Due to the very strong effects of x, w, and their interactions, the overall SD of y is `r rnd(sd(dcc$y))`.

So, the model coefficients here cannot be easily and directly interpreted in a correlation-like fashion.

Perhaps it's better to look at the actual standardized coefficients:

```{r}
lm.beta::lm.beta(fit1)
```


----------------------------------------------------------------------------------------

----------------------------------------------------------------------------------------

----------------------------------------------------------------------------------------

